### 2019-07-08 2-Chapters/1-Chapter/chapter1.tex
Done for proof reading intro FR/EN

\textbf{From IT to Cognitive Radio.}
%
We can narrow down the first field of study of this thesis, steps by steps:
from Information Technologies (IT), to Telecommunications, to Wireless Communications, then to Cognitive Radio (CR),
and finally to Software Defined Radio (SDR).
%
The transition from the historical approach of hardware-based radio to SDR architectures is a gradual process, that started in the early $1990$s and has accelerated in the $2000$s.
% https://en.wikipedia.org/wiki/Software-defined_radio
A SDR is a radio communication system where components that have been traditionally implemented in hardware (\eg, mixers, filters, modulators/demodulators, detectors, etc) are more and more implemented by means of software on a processor.
%  with the exception of a unique pair of receiver/transmitter antennas.
Even though the SDR paradigm was initiated by the United States of North America defense research in the $1980$s, over the past $20$ years the civil industry has started to be interested by SDR, and CR has got significant attention from the academia as well.
% CR has been the first answer to solve spectrum scarcity issue which collected a huge attention in early 21st century.
%
As CR is not a standard technology, it does not have a single definition, so let us start by quoting the definition of two researchers whose works constitutes the roots of the development of CR, the first one having appeared just $20$ years ago.
% - technologies de l'information > télécommunications > communications sans fil > radio intelligente > DSA > OSA
%
% je peux faire comme Navik, donner des définitions de la CR, citant d'autres travaux
\begin{itemize}\tightlist
    \item
    Joseph Mittola in $1999$ proposed that
    \emph{``a really smart radio that would be self-, RF- and user-aware, and that would include software technology and machine learning capabilities along with a lot of high-fidelity knowledge of the radio environment''} \cite{Mitola99}.

@@

\textbf{From Statistics or Machine Learning to Multi-Armed Bandits.}
%
% Similarly, we can narrow down the second field of study of this thesis, steps by steps.
% Stats > sequential learning > bandits
% A first point-of-view is the historical one, as
We are interested by Multi-Armed Bandits (MAB) learning,
which emerged as an interesting problem in the statistics community and among researchers interested by sequential learning, as shown by the pioneer works of \cite{Thompson33,Robbins52,LaiRobbins85}.
It is also included in the more general concept of Reinforcement Learning (RL), itself one of the field of Machine Learning (ML).
% ML > RL > MAB
% We can also expose another point-of-view, which starts from the general concept of Machine Learning (ML), and focus on , then on MAB learning.
% As illustrated in Figure~\ref{fig:1:MainTypesOfMachineLearning}, there are other types of ML, that we do not focus on in this thesis.
%
The reference book on RL is \cite{SuttonBarto2018}, so let us quote the definition of RL given by R. Sutton and A. Barto:
\emph{``Reinforcement learning is learning what to do - how to map situations to actions - so
as to maximize a numerical reward signal. The learner is not told which actions to
take, but instead must discover which actions yield the most reward by trying them''}.
% - apprentissage statistique (machine learning) > séquentiel > par renforcement > information partielle (bandits)

@@

\textbf{Opportunistic Spectrum Access.}
%
The two communities of OSA and MAB have started to interact, and the resulting works have received a great interest from both communities, since the late $2000$s and early $2010$s, with pioneers works like \cite{Liu08,Zhao10,Jouini09,Jouini10}.
The focus is on one SU accessing a licensed spectrum, occupied by PU that have a strict priority over the SU, which has to follow a ``listen-before-talk'' access scheme.
%
The following hypotheses are maid:
the SU is equipped with a spectrum sensing capability,
and considers a fixed and finite set of $K$ orthogonal channels, \ie, different frequency bands in a licensed spectrum.
For instance, it can be a set of $K=3$ Wi-Fi channels at different frequency, emitted by the same Wi-Fi station.
Another hypothesis is that both PU and SU are synchronized in time, by sub-dividing the time in discrete time steps (or rounds).
%
Thus if the SU spends a short time in the beginning of each time slot to perform sensing, it can scan for the presence or absence of any PU before transmitting, without colliding with the PU during the rest of the time slot.
If the SU was able to sense for all the $K$ frequency bands, it could simply transmit in one of the free channels if any, or not transmit if all channels are used at a given time step.
However, sensing is known to be costly, in terms of energy consumption, especially for wide-band sensing, as detailed in the surveys \cite{yucek2009survey,subhedar2011spectrum}, thus most works on OSA limit the sensing capacity of SU to sensing only one channel at a time.
This assumption, along with the hypothesis that PU cannot be disturbed (without which no regulation will ever be accepted for OSA), enforces the SU to transmit in the channel that it sensed, if and only if it was sensed free from any PU.
% non negligible cost of radio reconfiguration

@@

\textbf{MAB for OSA, and a short history of these researches at the SCEE team.}
%
Previous works of our SCEE team showed that MAB can be used to model the OSA problem:
orthogonal frequency bands (or channels) are modeled by arms $k\in\{1,\dots,K\}$,
and the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$ is modeled by a reward of $r(t) \in \{0,1\}$.
Indeed, $r(t) = 1$ indicates that no Primary User was sensed (and thus the SU can send a message), while a reward of $r(t)=0$ indicates that the channel $k$ is busy at during the time slot $t$, and no message should be sent, until the end of that slot.
%
This model was first studied by W. Jouini during his PhD thesis with C. Moy \cite{Jouini12PhD}, about ten years ago, first in the article \cite{Jouini09} and later in \cite{Jouini10,Jouini12}.

Their works were among the first ones to propose to use Reinforcement Learning for Cognitive Radio and the OSA problem, especially the MAB model and the $\UCB_1$ algorithm,
along with the early works of Q. Zhao and her team, for instance with \cite{Liu08,Zhao10}.
%
Shortly after in $2014$, proof-of-concepts using real-world radio hardware and Software Defined Radio were developed by C. Moy and his student C. Robert, using USRP boards and the MATLAB/Simulink software \cite{RobertSDR2014,MoyWSR2014}.
In a second PhD thesis, N. Modi studied from $2014$ to $2017$ the impact on the battery life of a wireless device of using MAB algorithms to optimize channel selections \cite{Modi17PhD}.
On the one hand, running a MAB algorithm such as UCB-like algorithms was proven to be useful and could bring significant improvement in terms of successful transmission rates, directly increasing the battery life of the device.
On the other hand, classical MAB algorithms tend to switch arms a lot of times, especially in the beginning of their learning process, and this induces a lot of dynamical hardware reconfigurations for the wireless device, as selecting a different channel requires a change in the radio hardware used by the device.
Each hardware reconfiguration costs energy for the device, and so quickly-switching algorithms lead to a reduction of the battery life.
The trade-off between the two aspects is studied empirically in
\cite{modiDemo2016}.

@@

\textbf{Limitations and specifities of IoT networks.}
%
The aforementioned literature have essentially shown that MAB algorithms can be applied successfully on the SU side for the OSA problem.
However, if we consider CR-ready devices that cannot perform spectrum sensing, such as low-cost and low energy-consumption end-devices designed for the future Internet of Things networks, the MAB model that uses sensing to detect PU in the OSA case can no longer be applied.
Moreover, in most cases, the IoT networks use unlicensed bands, and as such there is no longer a distinction between PU and SU.
In other words, there is no longer any priority between users in IoT networks.
%  (no more PU, all are SU).
%
The other specificities of IoT networks can be listed as follows,
and we refer to the survey \cite{Centenaro16} for more details.
\begin{itemize}\tightlist
    % \item
    % we consider here IoT networks that run in unlicensed bands (no more distinction between PU and SU),
    \item
    most IoT devices are very low-cost devices, run on low-quality hardware, and have limited software and power capabilities (\ie, tiny batteries).
    It means that even if they are not equipped with particular spectrum sensing capabilities,
    they are equipped with a transmitter as well as with a receiver (Tx and Rx, \ie, a transceiver),
    and they have low-complexity storage and computation capabilities (\ie, small embedded processors),
    \item
    a single IoT base station will have to handle a very large number of devices,
    and cannot send coordination orders to them in order to optimize the network,
    \item
    IoT devices have low to very low duty cycles (few messages every minute to only a few every month), and are in charge of initiating their up-link communications.
\end{itemize}

@@

% \paragraph{Chapter~\ref{chapter:2}.}
%
We started by exploring the rich literature of multi-armed bandits,
as many different algorithms exist, with lots of variants on the simple problem presented above.
% (see \cite{LattimoreBanditAlgorithmsBook,Slivkins2019} for surveys).
We thus start the first Part~\ref{part:Introduction} of this thesis with Chapter~\ref{chapter:2}, which presents the MAB model and gives a review of the most important algorithms designed to solve stochastic and stationary MAB problems.
%
In order to clearly understand which algorithms could be adapted to the aforementioned constraints of CR for IoT networks,
we were not only interested by the usual measure of performance of any MAB algorithm (its regret, see below in Section~\ref{sec:2:lowerUpperBoundsRegret}),
but also by their empirical performances in terms of computational complexity and storage requirements, from the point of view of both theoretical analyses and real-world measurements.
%  on time and memory footprints.

% \paragraph{Chapter~\ref{chapter:3}.}
%
Our exploration of the large number of MAB algorithms and models developed in the recent literature
has given us the ambition to write a single piece of software allowing anyone to easily implement new models and algorithms, in order to compare the existing ones and empirically explore the performances of newly proposed algorithms.
To tackle this goal, we developed a Python library of simulation of MAB problems.
%
To the best of our knowledge, we wrote the most exhaustive open-source library for bandits. Our library SMPyBandits is published online, under an open-source licence \cite{SMPyBanditsJMLR,SMPyBandits}, and hosted on \href{https://GitHub.com/SMPyBandits}{\texttt{GitHub.com/SMPyBandits}}.
We present in details its architecture and its features in Chapter~\ref{chapter:3}, along with different examples of its usage.
A full documentation is available online, as well as exhaustive instructions to reproduce the simulations presented in the rest of this thesis.

% \paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many MAB algorithms available, we are also interested by the question of how a practitioner can chose the algorithm to implement, in a given IoT object, in order to equip this object with the capacity to adapt robustly to any environment.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well known existing algorithms, in Section~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}.
We confirm that widely used and not too sophisticated algorithms, such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal}, are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is an online algorithm selection, consisting in aggregating a (finite) set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.
This work on aggregating MAB algorithms was motivated for the OSA case of Cognitive Radio, where many previous research works only considered the $\UCB_1$ algorithm without really justifying this choice.
This contribution was presented at the IEEE WCNC conference in Barcelona, Spain, in April $2018$.

@@

In the second Part~\ref{part:MABIOT} of this thesis, we start by proposing and studying different models of IoT networks, with simulations and a real-world proof-of-concept, and then we study interesting questions on two mathematical models of MAB, arising from simplification of our first model.

% \paragraph{Chapter~\ref{chapter:4}.}
%
We study two models of IoT networks in Chapter~\ref{chapter:4}.
The first model considers IoT devices that simply have data to regularly send to a base station (at unpredictable or random times).
Such devices use acknowledgements as feedback, in order to optimize their up-link communications, by accessing more the best channels (\ie, the channel less occupied by the surrounding traffic).
The goal of this first model is to increase the Quality of Service (QoS) of the application of such IoT network, by applying decentralized RL on the device side in order to reduce the rate of failed transmission.
The base station will thus receive more up-link packets from the devices being served, if they can successfully learn by themselves an efficient spectrum access scheme.
It implies that the network can accept more end-devices while maintaining the same QoS.
%
Our second model then considers packet retransmissions, and while the two concepts are similar, the application of an efficient decentralized learning algorithm now results in both an increased battery life for each of the devices, as well as an increased QoS for the entire network, as less retransmissions reduce the local spectrum load.

@@

The first model comes from the first article written during this thesis, which initiated a collaboration with R. Bonnefoi, another PhD student of our team SCEE.

@@

The proof-of-concept (PoC) continued the fruitful collaboration, and was demonstrated during three days at the IEEE ICT conference in Saint-Malo, France, in June $2018$ \cite{Besson2018ICT}, and later presented at the IEEE WCNC conference in Marrakech, Morocco, in April $2019$ \cite{Besson2019WCNC}.

@@

% \paragraph{Chapter~\ref{chapter:5}.}
%
In both models of IoT networks presented above, the core concept is to let every dynamic device run a learning algorithm in a fully decentralized way, in order to optimize the complete system.
It is important to note that each devices communicates, and thus learns, only at some rounds and not at all the time steps, by following its own random activation process (we restrict to a purely random Bernoulli activation process of probability $p$).
It means that each of them target its own local objective, which is to maximize its cumulated reward, in a selfish way and by being agnostic to the other devices following the same objective.
It is well known in game theory that playing selfishly can be disastrous for the centralized performance measure, one can think of popular ``dilemmas'', such as the \href{https://en.wikipedia.org/wiki/Prisoner%27s_dilemma}{prisoner dilemma}.
% , and for more examples we refer to the book \cite{CesaLugosi06}.
% https://www.ii.uni.wroc.pl/~lukstafi/pmwiki/uploads/AGT/Prediction_Learning_and_Games.pdf
So it was quite surprising that the numerical simulations, as well as the realistic PoC, showed that decentralized selfish MAB learning lead to efficient coordination between devices in all the considered scenarios, despite the fact that these IoT objects cannot communicate directly with each other, and receive only a collision feedback from the base station they are associated to.

That is why we first tried to analyze the performance of this decentralized algorithm, that we refer to as \Selfish, in the model of Section~\ref{sec:4:firstModel},
but due to the random number of active devices at each time step (\ie, as soon as the probability $p$ of activation is $p < 1$), we were unable to develop a clean analysis.
%
That is the reason why we relax the hypothesis of having $M \gg K$ (or even simply $M > K$) devices in a network with $K$ orthogonal wireless channel in Chapter~\ref{chapter:5}, that is equivalent to having $p < 1$, and we consider the case of IoT devices communicating at every time step (\ie, $p=1$), and so we restrict to only up-to $M \leq K$ devices.
The model we studied comes with different variants, depending on the feedback level, and covers both the OSA case (\ie, with sensing of PU) or the IoT case (\ie, without sensing).
The goal was to understand the heuristic of Chapter~\ref{chapter:4}, \Selfish, in the simpler framework of multi-player MAB, which has been studied before for the case of OSA, as studied for instance by \cite{Zhao10,Anandkumar10,Anandkumar11}.
%
The OSA case covered by the studied model can seem to slightly differ with the one considered by \cite{Jouini10} and other previous works,
as our model requires that an \Ack{} is sent back by the base station if the transmission was successful, even in the case where sensing information is available.
It is actually equivalent to the previously studied models of applying MAB for OSA, as they only considered synchronized devices, one SU and PU, and thus if the sensing indicates that one channel is free at one time slot, the up-link message sent by the SU device is sure to be successfully received by the base station (in the ideal model), so there is no risk of collision, and thus no need for an \Ack.

On the one hand, in the OSA setting, we failed to obtain positive result for the \Selfish{} policy, as we proved that in some limited settings (\eg, $K=3$ and $M=2$), \Selfish-\UCB{} can suffer from linear regret with a small probability, and thus suffers from linear mean regret.
This result was later confirmed and analyzed by other researchers in \cite{BoursierPerchet18} (Appendix~F).
%
On the other hand, we were able to propose new algorithms for this multi-player bandit problem with sensing information, and we analyzed our proposal \MCTopM-\klUCB, to show that it achieves a finite-time logarithmic regret upper bound, improving significantly over previous state-of-the-art results.
Our algorithm also achieves a logarithmic number of collisions and arm switches, and allows a fixed group of $M$ devices to efficiently learn to use the $M$ best channels orthogonally for almost all their up-link communications.
%
This strong theoretical result heavily depends on the presence of sensing feedback, and as such our results are not (yet) applicable to the IoT model without sensing.

@@

As explained above, the model of Chapter~\ref{chapter:4} was found intractable to analyze, mainly because we consider an IoT network with many end-devices, all following random activation patterns.
The difficulty does not reside in the fact that we are trying to analyze MAB algorithms, designed to tackle stationary problems, on a non-stationary problem,
but rather that we are analyzing algorithms which are all playing in different (random) activation times, subsets of the global synchronized time steps.
Generalizing to different activation probabilities would lead to a model even harder to analyze, and enforcing at most $M \leq K$ devices is in fact not really realistic for IoT networks, even though this lead to the interesting model of Chapter~\ref{chapter:5}.

On the one hand, if the activation pattern of all devices could be fixed, for instance based on a centralized affectation of the devices to different time slots, then the model of multi-player bandit from Chapter~\ref{chapter:5} can be used to let every group of $M \leq K$ devices learn an optimal orthogonal affectation in the $K$ channels (\eg, groups can be the set of devices that all emit at the same time).
%
However, even if we continue to assume a synchronized time in all the rest of the thesis,
it is hard to argue that this hypothesis of a centralized time schedule for the end-devices can be realistic, because we study the case of decentralized learning for OSA and IoT precisely in order to avoid signaling and any central control of the devices by the base station, as we explained above.

On the other hand, another way to relax the non-stationary hypothesis is to take the point-of-view of a single device, like in the OSA case mentioned above, where the focus is on one SU surrounded by many PU having stationary behaviors.
If we focus on one IoT device, its environment (\ie, the surrounding devices) is non-stationary, meaning that its average properties can fluctuate with time, but real environments usually present non-stationarities with certain structures.
For instance, enforcing stationarity on some time intervals leads to the last contribution and the last chapter of this thesis.
% DONE explain why is it interesting to study piece-wise stationary problems from the point-of-view of one player...

>>>>

\paragraph{From IT to Cognitive Radio.}
%
We can narrow down the first field of study of this thesis, steps by steps:
from Information Technologies (IT), to Telecommunications, to Wireless Communications, then to Cognitive Radio (CR),
and finally to Software Defined Radio (SDR).
%
The transition from the historical approach of hardware-based radio to SDR architectures is a gradual process, that started in the early $1990$s and has accelerated in the $2000$s.
% https://en.wikipedia.org/wiki/Software-defined_radio
A SDR is a radio communication system where components that have been traditionally implemented in hardware (\eg, mixers, filters, modulators/demodulators, detectors, etc) are more and more implemented by means of software on a processor.
%  with the exception of a unique pair of receiver/transmitter antennas.
Even though the SDR paradigm was initiated by the United States of North America defense research in the $1980$s, over the past $20$ years the civil industry has started to be interested by SDR, and CR has got significant attention from the academia as well.
% CR has been the first answer to solve spectrum scarcity issue which collected a huge attention in early 21st century.
%
As CR is not a standard technology, it does not have a single definition, so let us start by quoting the definition of two researchers whose works constitutes the roots of the development of CR, the first one having appeared just $20$ years ago.
% - technologies de l'information > télécommunications > communications sans fil > radio intelligente > DSA > OSA
%
% je peux faire comme Navik, donner des définitions de la CR, citant d'autres travaux
\begin{itemize}\tightlist
    \item
    Joseph Mittola in $1999$ proposed that
    \emph{``a really smart radio that would be self-, RF- and user-aware, and that would include software technology and machine learning capabilities along with a lot of high-fidelity knowledge of the radio environment''} \cite{Mitola99}.

@@

\paragraph{From Statistics or Machine Learning to Multi-Armed Bandits.}
%
% Similarly, we can narrow down the second field of study of this thesis, steps by steps.
% Stats > sequential learning > bandits
% A first point-of-view is the historical one, as
We are interested by Multi-Armed Bandits (MAB) learning,
which emerged as an interesting problem in the statistics community and among researchers interested by sequential learning, as shown by the pioneer works of \cite{Thompson33,Robbins52,LaiRobbins85}.
It is also included in the more general concept of Reinforcement Learning (RL), itself one of the field of Machine Learning (ML).
% ML > RL > MAB
% We can also expose another point-of-view, which starts from the general concept of Machine Learning (ML), and focus on , then on MAB learning.
% As illustrated in Figure~\ref{fig:1:MainTypesOfMachineLearning}, there are other types of ML, that we do not focus on in this thesis.
%
The reference book on RL is \cite{SuttonBarto2018}, so let us quote the definition of RL given by R. Sutton and A. Barto:
\emph{``Reinforcement learning is learning what to do - how to map situations to actions - so
as to maximize a numerical reward signal. The learner is not told which actions to
take, but instead must discover which actions yield the most reward by trying them''}.
% - apprentissage statistique (machine learning) > séquentiel > par renforcement > information partielle (bandits)

@@

\paragraph{Opportunistic Spectrum Access.}
%
The two communities of OSA and MAB have started to interact, and the resulting works have received a great interest from both communities, since the late $2000$s and early $2010$s, with pioneers works like \cite{Liu08,Zhao10,Jouini09,Jouini10}.
The focus is on one SU accessing a licensed spectrum, occupied by PU that have a strict priority over the SU, which has to follow a ``listen-before-talk'' access scheme.
%
The following hypotheses are maid:
the SU is equipped with a spectrum sensing capability,
and considers a fixed and finite set of $K$ orthogonal channels, \ie, different frequency bands in a licensed spectrum.
For instance, it can be a set of $K=3$ Wi-Fi channels at different frequency, emitted by the same Wi-Fi station.
Another hypothesis is that both PU and SU are synchronized in time, by sub-dividing the time in discrete time steps (or rounds).
%
Thus if the SU spends a short time in the beginning of each time slot to perform sensing, it can scan for the presence or absence of any PU before transmitting, without colliding with the PU during the rest of the time slot.
If the SU was able to sense for all the $K$ frequency bands, it could simply transmit in one of the free channels if any, or not transmit if all channels are used at a given time step.
However, sensing is known to be costly, in terms of energy consumption, especially for wide-band sensing, as detailed in the surveys \cite{yucek2009survey,subhedar2011spectrum}, thus most works on OSA limit the sensing capacity of SU to sensing only one channel at a time.
This assumption, along with the hypothesis that PU cannot be disturbed (without which no regulation will ever be accepted for OSA), enforces the SU to transmit in the channel that it sensed, if and only if it was sensed free from any PU.
% non negligible cost of radio reconfiguration

@@

\paragraph{MAB for OSA, and a short history of these researches at the SCEE team.}
%
Previous works of our SCEE team showed that MAB can be used to model the OSA problem:
orthogonal frequency bands (or channels) are modeled by arms $k\in\{1,\dots,K\}$,
and the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$ is modeled by a reward of $r(t) \in \{0,1\}$.
Indeed, $r(t) = 1$ indicates that no Primary User was sensed (and thus the SU can send a message), while a reward of $r(t)=0$ indicates that the channel $k$ is busy at during the time slot $t$, and no message should be sent, until the end of that slot.
%
This model was first studied by W. Jouini during his PhD thesis with C. Moy \cite{Jouini12PhD}, about ten years ago, first in the article \cite{Jouini09} and later in \cite{Jouini10,Jouini12}.

Their works were among the first ones to propose to use Reinforcement Learning for Cognitive Radio and the OSA problem, especially the MAB model and the $\UCB_1$ algorithm,
along with the early works of Q. Zhao and her team, for instance with \cite{Liu08,Zhao10}.
%
Shortly after in $2014$, proof-of-concepts using real-world radio hardware and Software Defined Radio were developed by C. Moy and his student C. Robert, using USRP boards and the MATLAB/Simulink software \cite{RobertSDR2014,MoyWSR2014}.
In a second PhD thesis, N. Modi studied from $2014$ to $2017$ the impact on the battery life of a wireless device of using MAB algorithms to optimize channel selections \cite{Modi17PhD}.
% On the one hand, running a MAB algorithm such as UCB-like algorithms was proven to be useful and could bring significant improvement in terms of successful transmission rates, directly increasing the battery life of the device.
% On the other hand, classical MAB algorithms tend to switch arms a lot of times, especially in the beginning of their learning process, and this induces a lot of dynamical hardware reconfigurations for the wireless device, as selecting a different channel requires a change in the radio hardware used by the device.
% Each hardware reconfiguration costs energy for the device, and so quickly-switching algorithms lead to a reduction of the battery life.
The trade-off between rapid software reconfigurations that cost energy but allow to learn quickly is studied empirically in \cite{modiDemo2016}.

@@

\paragraph{Limitations and specifities of IoT networks.}
%
The aforementioned literature have essentially shown that MAB algorithms can be applied successfully on the SU side for the OSA problem.
However, if we consider CR-ready devices that cannot perform spectrum sensing, such as low-cost and low energy-consumption end-devices designed for the future Internet of Things networks, the MAB model that uses sensing to detect PU in the OSA case can no longer be applied.
Moreover, in most cases, the IoT networks use unlicensed bands, and as such there is no longer a distinction between PU and SU.
In other words, there is no longer any priority between users in IoT networks.
%  (no more PU, all are SU).
%
The other specificities of IoT networks can be listed as follows,
and we refer to the survey \cite{Centenaro16} for more details.
\begin{itemize}\tightlist
    % \item
    % we consider here IoT networks that run in unlicensed bands (no more distinction between PU and SU),
    \item
    most IoT devices are very low-cost devices, run on low-quality hardware, and have limited software and power capabilities (\ie, tiny batteries).
    It means that even if they are not equipped with particular spectrum sensing capabilities,
    they are equipped with a transmitter as well as with a receiver (Tx and Rx, \ie, a transceiver),
    and they have low-complexity storage and computation capabilities (\ie, small embedded processors),
    \item
    a single IoT base station will have to handle a very large number of devices,
    and cannot send coordination orders to them in order to optimize the network,
    \item
    IoT devices have low to very low duty cycles (few messages every minute to only a few every month), and are in charge of initiating their up-link communications.
\end{itemize}

@@

% \paragraph{Chapter~\ref{chapter:2}.}
%
We started by exploring the rich literature of multi-armed bandits,
as many different algorithms exist, with lots of variants on the simple problem presented above.
% (see \cite{LattimoreBanditAlgorithmsBook,Slivkins2019} for surveys).
We thus start the first part of this thesis with Chapter~\ref{chapter:2}, which presents the MAB model and reviews the most important algorithms designed to solve stochastic and stationary MAB problems.
%
In order to clearly understand which algorithms could be adapted to the aforementioned constraints of CR for IoT networks,
we were not only interested by the usual measure of performance of any MAB algorithm (its regret, see below in Section~\ref{sec:2:lowerUpperBoundsRegret}),
but also by their empirical performances in terms of computational complexity and storage requirements, from the points of view of both theoretical analyses and real-world measurements.
%  on time and memory footprints.

% \paragraph{Chapter~\ref{chapter:3}.}
%
Our exploration of the large number of MAB algorithms and models developed in the recent literature
has given us the ambition to write a single software allowing anyone to easily implement new models and algorithms, in order to compare the existing ones and empirically explore the performances of newly proposed algorithms.
To tackle this goal, we developed a library of simulation of MAB problems, in the Python language.
%
To the best of our knowledge, we wrote the most exhaustive open-source library for bandits. Our library SMPyBandits is published online, under an open-source licence \cite{SMPyBanditsJMLR,SMPyBandits}, and hosted on \href{https://GitHub.com/SMPyBandits}{\texttt{GitHub.com/SMPyBandits}}.
We present in details its architecture and its features in Chapter~\ref{chapter:3}, along with different examples of its usage.
A full documentation is available online, as well as exhaustive instructions to reproduce the simulations presented in the rest of this thesis.

% \paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many MAB algorithms available, we are also interested by the question of how an engineer can chose the algorithm to implement, in a given IoT object, in order to equip this object with the capacity to adapt robustly to any environment.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well known existing algorithms, in Section~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}.
We confirm that widely used and not too sophisticated algorithms, such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal}, are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is online algorithm selection, consisting in aggregating a (finite) set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.
This work on aggregating MAB algorithms was motivated for the OSA case of CR, where many previous research works only considered the $\UCB_1$ algorithm without really justifying their choice.
This contribution was presented at the IEEE WCNC conference in Barcelona, Spain, in April $2018$.

@@

In the second part of this thesis, we start by proposing and studying different models of IoT networks, with simulations and a real-world proof-of-concept, and then we study interesting questions on two mathematical models of MAB, arising from simplification of our first model.

% \paragraph{Chapter~\ref{chapter:4}.}
%
We study two models of IoT networks in Chapter~\ref{chapter:4}.
The first model considers IoT devices that simply have data to regularly send to a base station (at unpredictable or random times).
Such devices use acknowledgements as feedback, in order to optimize their up-link communications, by accessing more the best channels (\ie, the channel less occupied by the surrounding traffic).
The goal of this first model is to increase the Quality of Service (QoS) of such IoT network, by applying decentralized RL on the device side in order to reduce the rate of failed transmissions.
The base station will thus receive more up-link packets from the devices being served, if they can successfully learn by themselves an efficient spectrum access scheme.
It implies that the network can serve more end-devices while maintaining the same QoS.
%
Our second model then considers packet retransmissions, and while the two models are similar, the application of an efficient decentralized learning algorithm now results in both an increased battery life for each of the devices, as well as an increased QoS for the entire network, as less retransmissions reduce the local spectrum load.

@@

The model without retransmission comes from the first article written during this thesis, which initiated a collaboration with R. Bonnefoi, another PhD student of our team SCEE.

@@

The proof-of-concept (PoC) continued the fruitful collaboration, and was demonstrated during three days at the IEEE ICT conference in Saint-Malo, France, in June $2018$ \cite{Besson2018ICT}, and later at the IEEE WCNC conference in Marrakech, Morocco, in April $2019$ \cite{Besson2019WCNC}.

@@

% \paragraph{Chapter~\ref{chapter:5}.}
%
In both models of IoT networks presented above, the core idea is to let every dynamic device run a learning algorithm in a fully decentralized way, in order to optimize the entire system.
It is important to note that each devices communicates, and thus learns, only at some rounds and not at all the time steps, by following its own random activation process (we restrict to a purely random Bernoulli activation process of probability $p$).
It means that each of them target its own local objective, which is to maximize its cumulated reward, in a selfish way and by being agnostic about the other devices following the same objective.
It is well known in game theory that playing selfishly can be disastrous for the centralized performance measure, one can think of popular ``dilemmas'', such as the \href{https://en.wikipedia.org/wiki/Prisoner%27s_dilemma}{prisoner dilemma}.
% , and for more examples we refer to the book \cite{CesaLugosi06}.
% https://www.ii.uni.wroc.pl/~lukstafi/pmwiki/uploads/AGT/Prediction_Learning_and_Games.pdf
So it was quite surprising that the numerical simulations, as well as the realistic PoC, showed that decentralized selfish MAB learning lead to efficient coordination between devices in all the considered scenarios, despite the fact that these IoT objects cannot communicate directly with each other, and receive only a collision feedback from the base station they are associated to.

That is why we first tried to analyze the performance of this decentralized algorithm, that we refer to as \Selfish, in the model of Section~\ref{sec:4:firstModel},
but due to the random number of active devices at each time step (\ie, as soon as the probability $p$ of activation is $p < 1$), we were unable to develop a clean analysis.
%
That is the reason why we relax the hypothesis of having $M \gg K$ (or even simply $M > K$) devices in a network with $K$ orthogonal wireless channel in Chapter~\ref{chapter:5}, that is equivalent to having $p < 1$, and we consider the case of IoT devices communicating at every time step (\ie, $p=1$), and so we restrict to only up-to $M \leq K$ devices.
The model we studied comes with different variants, depending on the feedback level, and covers both the OSA case (\ie, with sensing of PU) or the IoT case (\ie, without sensing).
The goal was to understand the heuristic of Chapter~\ref{chapter:4}, \Selfish, in the simpler framework of multi-player MAB, which has been studied before for the case of OSA, for instance by \cite{Zhao10,Anandkumar10,Anandkumar11}.
%
The OSA case covered by the studied model can seem to slightly differ with the one exposed above, from \cite{Jouini10} and other previous works,
as our model requires that an \Ack{} is sent back by the base station if the transmission was successful, even in the case where sensing information is available.
It is actually equivalent to the previously studied models of applying MAB for OSA, as they only considered synchronized devices, only one SU and PU, and thus if the sensing indicates that one channel is free at one time slot, the up-link message sent by the SU device is sure to be successfully received by the base station (in the ideal model), so there is no risk of collision, and thus no need for an \Ack.

On the one hand, in the OSA setting, we failed to obtain positive result for the \Selfish{} policy, as we proved that in some limited settings (\eg, $K=3$ channels and $M=2$ devices), \Selfish-\UCB{} can suffer from linear regret with a small probability, and thus suffers from linear mean regret.
This result was later confirmed and analyzed by other researchers in \cite{BoursierPerchet18} (Appendix~F).
%
On the other hand, we were able to propose new algorithms for this multi-player bandit problem with sensing information, and we analyzed our proposal \MCTopM{} to show that it achieves a finite-time logarithmic regret upper bound, and improves significantly over previous state-of-the-art results.
Our algorithm also achieves a logarithmic number of collisions and arm switches, and allows a fixed group of $M$ devices to efficiently learn to use the $M$ best channels orthogonally for almost all their up-link communications.
%
This strong theoretical result heavily depends on the presence of sensing feedback, and as such our results are not (yet) applicable to the IoT model without sensing.

@@

As explained above, the model of Chapter~\ref{chapter:4} was found intractable to analyze, mainly because we consider an IoT network with many end-devices, all following random activation patterns.
The difficulty does not reside in the fact that we are trying to analyze MAB algorithms, designed to tackle stationary problems, applied on a non-stationary problem,
but rather that we consider algorithms which are all playing in different (random) activation times, subsets of the global synchronized time steps.
Generalizing to different activation probabilities would lead to a model even harder to analyze, and enforcing at most $M \leq K$ devices is in fact not really realistic for IoT networks, even though this lead to the interesting model of Chapter~\ref{chapter:5}.

On the one hand, if the activation pattern of all devices could be fixed, for instance based on a centralized affectation of the devices to different time slots, then the model of multi-player bandit from Chapter~\ref{chapter:5} can be used to let every group of $M \leq K$ devices learn an optimal orthogonal affectation in the $K$ channels (\eg, groups can be the set of devices that all emit at the same time).
%
However, even if we continue to assume a synchronized time in all the rest of the thesis,
it is hard to argue that this hypothesis of a centralized time schedule for the end-devices can be realistic, because we study the case of decentralized learning for OSA and IoT precisely in order to avoid signaling and any central control of the devices by the base station, as we explained before.

On the other hand, another way to relax the non-stationary hypothesis is to take the point-of-view of a single device, like in the OSA case mentioned above, where the focus is on one SU surrounded by many PU having stationary behaviors.
If we focus on one IoT device, its environment (\ie, the surrounding devices) is non-stationary, meaning that its average properties can fluctuate with time, but real environments usually present non-stationarities with certain structures.
%
It is thus interesting and realistic to enforce stationarity on some time intervals, and this leads to the last contribution and the last chapter of this thesis.
% DONE explain why is it interesting to study piece-wise stationary problems from the point-of-view of one player...
